---
title: "How to use this site"
description: "Choosing R or Python, running the examples yourself, the callout boxes, how results are reported, and how exercises work."
engine: knitr
---

```{r}
#| include: false
source("R/check_agree.R")
```

## Pick your language once

Every example on this site is written twice: once in **R** and once in **Python**. Both give the same answer. Click a tab to choose. The site remembers your choice on every page, so you only pick once.

Here is a tiny example: five pretend patients' ages, their mean, and their standard deviation (SD).

::: {.panel-tabset group="language"}
## R

```{r}
ages <- c(64, 71, 58, 69, 75)   # five pretend patients' ages

mean_age <- mean(ages)           # the average
sd_age <- sd(ages)               # the standard deviation

mean_age
sd_age
```

## Python

```{python}
import pandas as pd

ages = pd.Series([64, 71, 58, 69, 75])   # five pretend patients' ages

mean_age = ages.mean()                    # the average
sd_age = ages.std()                       # the standard deviation

print(mean_age)
print(sd_age)
```
:::

```{r}
#| include: false
check_agree(
  list(mean = mean_age, sd = sd_age),
  list(mean = reticulate::py$mean_age, sd = reticulate::py$sd_age)
)
```

Both say the mean age is 67.4 years with an SD of 6.6 years.

::: {.callout-tip}
## 🔀 R vs Python: the standard deviation trap
Python has two common ways to compute an SD, and they disagree. `pandas` (`ages.std()`) divides by *n − 1*, like R's `sd()`. `numpy` (`np.std(ages)`) divides by *n* unless you write `np.std(ages, ddof=1)`. Medical papers report the *n − 1* version. Whenever the two languages' defaults differ, this site sets them explicitly and explains why in a box like this one.
:::

Behind the scenes, every page checks that the R and Python answers match. If they ever disagree, the site refuses to build, so you can trust that both tabs say the same thing.

## Running the code yourself

1. Open the repository folder in Positron (see [Install & set up](setup.qmd)).
2. Save practice files in the `scratch/` folder, for example `scratch/practice.R` or `scratch/practice.py`. Git ignores everything in `scratch/`, so practice files can never be committed by accident.
3. Copy code with the copy button that appears in the top-right corner of every code block.
4. Run a line with **Cmd+Enter** (Mac) or **Ctrl+Enter** (Windows). Positron sends it to the console.

Data paths such as `data/cohort.csv` are relative to the repository's top folder. Positron starts there when you open the folder, so the paths just work.

## The boxes you'll see

::: {.callout-note}
## 💡 In plain language
The idea without the jargon. If a page loses you, read these boxes first.
:::

::: {.callout-warning}
## ⚠️ Watch out
Mistakes we see often, and how to avoid them.
:::

::: {.callout-tip}
## 🔀 R vs Python
Where the two languages' defaults differ, and what this site does about it.
:::

::: {.callout-note collapse="true"}
## 🔍 Under the hood
The math, for the curious. These start closed: click to open. You never need them to use a test correctly.
:::

## How results are reported

Every "How to report it" paragraph on this site follows the same rules, which match what JOA and JBJS reviewers expect.

| What | Report it as | Example |
|------|------|------|
| Continuous, roughly symmetric | mean (SD) | age 67.4 (6.6) years |
| Continuous, skewed | median (IQR) | length of stay 1 (0–2) days |
| Categories | n (%) | 312 (52%) female |
| p-values | three decimals; never "p = 0.000" or "p = 1.000" | p = 0.017; p < 0.001; p > 0.999 |
| Effect estimates | with a 95% confidence interval | difference 1.6 kg/m² (95% CI 0.7 to 2.5) |

The examples in the table are illustrations, not results from our data. [Demographics (Table 1)](../foundations/02-demographics.qmd) shows the full Table 1 layout.

**Results sentences give the numbers and stop there:** each group's value, the estimate with its 95% CI, and the p-value. Leave out words that interpret them, such as "similar", "higher than", "more often", "improved", "significant", "no clear evidence", or labels like "good" or "moderate". Saying what the numbers mean is the Discussion's job. An illustration:

> KOOS JR at 1 year was a median of 87 points (IQR 76 to 94) in men and 85 (IQR 73 to 94) in women (Hodges-Lehmann difference 1 point, 95% CI −2 to 4; p = 0.432).

Around each Results sentence, this site's text does interpret the numbers, because that's how you learn to read an output. The Results sentence itself is the part to copy into a manuscript.

## Exercises

Each page ends with two to four short exercises on the practice data. They get harder in three steps: run the same analysis on a different variable, pick the right test for a short scenario, then write the Results sentence. Solutions are hidden in both languages. Try first, then check.

**Try it:** using the same five ages, find the **median**.

::: {.callout-tip collapse="true"}
## Solution

::: {.panel-tabset group="language"}
## R

```r
ages <- c(64, 71, 58, 69, 75)
median(ages)   # 69
```

## Python

```python
import pandas as pd

ages = pd.Series([64, 71, 58, 69, 75])
ages.median()   # 69.0
```
:::

:::

## Getting help

- Something on this site is wrong or confusing? [Open an issue](https://github.com/hfr-ortho/stats-formation/issues) on GitHub.
- A question about your own project? Ask your TJS project lead.
