---
title: "18 · Example study report"
description: "A complete mini-study, from the raw abstraction workbook to a manuscript's Methods and Results, with a Word version to download."
engine: knitr
format:
  html: default
  docx:
    title: "Improvement in joint-specific scores one year after hip and knee replacement: a two-site cohort study"
    subtitle: "Synthetic data for teaching. Every patient in this report is invented."
---

```{r}
#| include: false
source("R/check_agree.R")
```

```{r}
#| echo: false
#| output: asis
# Stamp the data version into the page; tests/site/test_freshness.py checks it.
cat(sprintf("<!-- data-checksum: %s -->", unname(tools::md5sum("data/CHECKSUMS.md5"))))
```

::: {.content-visible when-format="html"}
This page puts the whole site together in one small study, from the messy abstraction workbook to the Methods and Results of a manuscript. Steps 1 to 7 do the work, in R and Python. Step 8 is the manuscript itself, and it's also a Word document: **MS Word**, under "Other Formats" in the right margin. The Word file contains only the manuscript, built from the same code every time the site is rendered.

::: {.callout-note}
## 💡 How this page works
Run the code blocks in order, from the top: later blocks use the packages and data loaded in the first one. Each step links to the page that explains its method.
:::

## 1 · The research question {#question}

Write the question down before you look at the outcome data, together with the primary outcome and the analysis that will answer it. Deciding these after seeing the results makes any p-value meaningless ([page 15](../beyond/15-post-hoc.qmd#planned-comparisons)). For this study:

- **Question:** do patients improve as much in the first year after a total knee replacement (TKA) as after a total hip replacement (THA)?
- **Primary outcome:** the change in the joint-specific score from before surgery to 1 year: HOOS JR after THA, KOOS JR after TKA, each scored 0 to 100, higher is better.
- **Primary analysis:** the difference in mean change between THA and TKA, with its 95% CI, from Welch's t test ([page 6](../catalog/06-two-unpaired-groups.qmd#unpaired-t)), if the changes look roughly symmetric within each group ([page 3](../foundations/03-distributions.qmd)). Otherwise the Mann-Whitney test.
- **Secondary outcome:** revision of any component, as a cumulative incidence with death as a competing risk ([page 13](../survival/13-kaplan-meier.qmd#competing-risks)).

::: {.callout-warning}
## ⚠️ Watch out: two questionnaires
HOOS JR and KOOS JR are different questionnaires. Both run from 0 to 100, but a 10-point gain on one isn't guaranteed to mean the same as a 10-point gain on the other. The comparison is still common in the literature, but say plainly in the Methods that it compares points on each joint's own scale. A stronger design uses a score every patient completes, such as the VR-12 physical component score.
:::

## 2 · Tidy the raw workbook {#tidy}

The data come from the abstraction workbook that [page 1](../foundations/01-tidy-data.qmd#tidying) tidies step by step: 60 cases from each site, with banner rows, merged headers, mixed dates, five spellings of "missing" and revisions marked only by a red fill. The code below is page 1's five steps in one block; open it to see the code, and go back to page 1 for why each step is there.

::: {.panel-tabset group="language"}
## R

```{r}
#| warning: false
library(tidyverse)
library(readxl)
library(tidyxl)
library(gtsummary)
library(survival)
library(ggsurvfit)

path <- "data/messy_abstraction_workbook.xlsx"
```

```{r}
#| code-fold: true
#| code-summary: "The tidying code from page 1"
# Step 1: read both tabs as text; the group header says pre-op first, then 1 year
read_site <- function(sheet) {
  raw <- read_excel(path, sheet = sheet, skip = 3, col_types = "text", .name_repair = "minimal")
  names(raw) <- str_trim(names(raw))
  names(raw)[names(raw) == "Date"]  <- c("prom_preop_date", "prom_1yr_date")
  names(raw)[names(raw) == "Score"] <- c("prom_preop", "prom_1yr")
  raw |>
    janitor::clean_names() |>
    mutate(site = sheet)
}
raw <- bind_rows(read_site("Site A"), read_site("Site B"))   # stack the tabs by column name

# Steps 2 and 3: drop the totals rows, duplicates and identifiers; one way to say "missing"
missing_codes <- c("N/A", "unk", "-", "999")
step3 <- raw |>
  filter(is.na(pt_name) | pt_name != "TOTAL") |>
  distinct() |>
  select(-pt_name, -mrn, -mua) |>
  mutate(across(everything(), \(x) if_else(x %in% missing_codes, NA, x)))

# Step 4: revisions are marked only by a red fill on the Study ID cell
cells <- xlsx_cells(path)
fills <- xlsx_formats(path)$local$fill$patternFill$fgColor$rgb
revised_ids <- cells |>
  filter(fills[local_format_id] %in% "FFFF9999") |>
  pull(character) |>
  str_trim() |>
  unique()

# Step 5: give every column its real type
parse_messy_date <- function(x, day_first = TRUE) {
  serial <- !is.na(x) & str_detect(x, "^\\d{5}$")
  iso    <- !is.na(x) & str_detect(x, "^\\d{4}-\\d{2}-\\d{2}$")
  rest   <- !is.na(x) & !serial & !iso
  out <- as.Date(rep(NA, length(x)))
  out[serial] <- janitor::excel_numeric_to_date(as.numeric(x[serial]))
  out[iso]    <- ymd(x[iso])
  out[rest]   <- if (day_first) dmy(x[rest], quiet = TRUE) else mdy(x[rest], quiet = TRUE)
  out
}

tidy <- step3 |>
  rename(procedure_text = procedure) |>
  transmute(
    case_id      = str_trim(study_id),
    site,
    surgery_date = parse_messy_date(dos),
    age          = parse_number(age),
    sex          = case_when(str_detect(sex, regex("^f", ignore_case = TRUE)) ~ "Female",
                             str_detect(sex, regex("^m", ignore_case = TRUE)) ~ "Male"),
    bmi          = parse_number(bmi),
    asa          = as.integer(as.roman(str_remove(asa, "^ASA "))),
    diabetes     = as.integer(str_detect(comorbidities, regex("\\bDM\\b", ignore_case = TRUE))),
    hypertension = as.integer(str_detect(comorbidities, regex("\\bHTN\\b", ignore_case = TRUE))),
    sleep_apnea  = as.integer(str_detect(comorbidities, regex("\\bOSA\\b", ignore_case = TRUE))),
    procedure    = case_when(str_detect(procedure_text, regex("TKA|knee", ignore_case = TRUE)) ~ "TKA",
                             str_detect(procedure_text, regex("THA|hip", ignore_case = TRUE)) ~ "THA"),
    side         = case_when(str_detect(procedure_text, "\\b(L|Left)\\b") ~ "L",
                             str_detect(procedure_text, "\\b(R|Right)\\b") ~ "R"),
    los_days     = parse_number(los),
    revised      = as.integer(case_id %in% revised_ids),
    prom_preop_date = parse_messy_date(prom_preop_date),
    prom_preop      = parse_number(prom_preop),
    prom_1yr_date   = parse_messy_date(prom_1yr_date),
    prom_1yr        = parse_number(prom_1yr)
  ) |>
  arrange(site, case_id)

dim(tidy)
```

## Python

```{python}
import re
import numpy as np
import pandas as pd
import openpyxl
import matplotlib.pyplot as plt
import pingouin as pg
import statsmodels.formula.api as smf
from scipy import stats
from scipy.stats import norm
from tableone import TableOne
from lifelines import KaplanMeierFitter
from lifelines.utils import qth_survival_times
from statsmodels.duration.survfunc import CumIncidenceRight

path = "data/messy_abstraction_workbook.xlsx"
```

```{python}
#| code-fold: true
#| code-summary: "The tidying code from page 1"
# Step 1: read both tabs as text; pandas names the repeated headers "Date.1" and "Score.1"
def read_site(sheet):
    raw = pd.read_excel(path, sheet_name=sheet, header=3, dtype=str)
    names = [str(c).strip() for c in raw.columns]
    pairs = {"Date": "prom_preop_date", "Score": "prom_preop",
             "Date.1": "prom_1yr_date", "Score.1": "prom_1yr"}
    names = [pairs.get(n, n) for n in names]
    raw.columns = [re.sub(r"[^0-9a-z]+", "_", n.lower()).strip("_") for n in names]
    return raw.assign(site=sheet)

raw = pd.concat([read_site("Site A"), read_site("Site B")], ignore_index=True)   # stack by column name

# Steps 2 and 3: drop the totals rows, duplicates and identifiers; one way to say "missing"
step3 = (raw[raw["pt_name"] != "TOTAL"]
         .drop_duplicates()
         .drop(columns=["pt_name", "mrn", "mua"])
         .replace(["N/A", "unk", "-", "999"], pd.NA))

# Step 4: revisions are marked only by a red fill on the Study ID cell
revised_ids = set()
for sheet in openpyxl.load_workbook(path).worksheets:
    for row in sheet.iter_rows(min_row=5):
        for cell in row:
            if cell.fill.fgColor.rgb == "FFFF9999":
                revised_ids.add(str(cell.value).strip())

# Step 5: give every column its real type
def number(text):
    return pd.to_numeric(text.str.extract(r"(\d+\.?\d*)")[0])

def parse_messy_date(col, dayfirst=True):
    iso = col.str.match(r"\d{4}-\d{2}-\d{2}", na=False)
    swiss = col.notna() & ~iso
    dates = pd.concat([pd.to_datetime(col[iso], format="ISO8601"),
                       pd.to_datetime(col[swiss], format="mixed", dayfirst=dayfirst)])
    return dates.reindex(col.index)

asa_roman = {"I": "1", "II": "2", "III": "3", "IV": "4"}
procedure_text = step3["procedure"]

tidy = pd.DataFrame({
    "case_id":      step3["study_id"].str.strip(),
    "site":         step3["site"],
    "surgery_date": parse_messy_date(step3["dos"]),
    "age":          number(step3["age"]),
    "sex":          step3["sex"].str[0].str.lower().map({"f": "Female", "m": "Male"}),
    "bmi":          number(step3["bmi"]),
    "asa":          pd.to_numeric(step3["asa"].str.replace("ASA ", "").replace(asa_roman)),
    "diabetes":     step3["comorbidities"].str.contains(r"\bDM\b", case=False).astype(int),
    "hypertension": step3["comorbidities"].str.contains(r"\bHTN\b", case=False).astype(int),
    "sleep_apnea":  step3["comorbidities"].str.contains(r"\bOSA\b", case=False).astype(int),
    "procedure":    np.select([procedure_text.str.contains(r"TKA|knee", case=False),
                               procedure_text.str.contains(r"THA|hip", case=False)],
                              ["TKA", "THA"], default=None),
    "side":         np.select([procedure_text.str.contains(r"\b(?:L|Left)\b"),
                               procedure_text.str.contains(r"\b(?:R|Right)\b")],
                              ["L", "R"], default=None),
    "los_days":     number(step3["los"]),
    "revised":      step3["study_id"].str.strip().isin(revised_ids).astype(int),
    "prom_preop_date": parse_messy_date(step3["prom_preop_date"]),
    "prom_preop":      number(step3["prom_preop"]),
    "prom_1yr_date":   parse_messy_date(step3["prom_1yr_date"]),
    "prom_1yr":        number(step3["prom_1yr"]),
}).sort_values(["site", "case_id"]).reset_index(drop=True)

print(tidy.shape)
```
:::

```{r}
#| include: false
# Spec 8.4: the tidying must reproduce the answer key exactly, as on page 1.
key <- read_csv("data/answer-keys/abstraction_workbook_tidy.csv", show_col_types = FALSE)
stopifnot(isTRUE(all.equal(as.data.frame(tidy), as.data.frame(key), check.attributes = FALSE)))
```

```{python}
#| include: false
key = pd.read_csv("data/answer-keys/abstraction_workbook_tidy.csv",
                  parse_dates=["surgery_date", "prom_preop_date", "prom_1yr_date"])
pd.testing.assert_frame_equal(tidy, key, check_dtype=False)
```

120 cases, 18 columns. The workbook has no follow-up: it records whether a hip or knee was revised, but not when, and not whether the patient died. Follow-up comes from the joint registry, linked by study ID. In the practice data, `data/cohort.csv` plays the registry. Take only the columns you need from it:

::: {.panel-tabset group="language"}
## R

```{r}
registry <- read_csv("data/cohort.csv", show_col_types = FALSE) |>
  select(case_id, patient_id, followup_years, event_status, revised_registry = revised,
         surgery_date_registry = surgery_date, procedure_registry = procedure)   # these three, to check the link

study <- left_join(tidy, registry, by = "case_id")   # keeps every workbook case
nrow(study)
count(study, procedure)
```

## Python

```{python}
registry = (pd.read_csv("data/cohort.csv", parse_dates=["surgery_date"])
            [["case_id", "patient_id", "followup_years", "event_status", "revised", "surgery_date", "procedure"]]
            .rename(columns={"revised": "revised_registry", "surgery_date": "surgery_date_registry",
                             "procedure": "procedure_registry"}))   # these three, to check the link

study = tidy.merge(registry, on="case_id", how="left")   # keeps every workbook case
print(len(study))
print(study["procedure"].value_counts().sort_index().to_string())
```
:::

```{python}
#| include: false
chk = {"rows": float(len(study)), "patients": float(study["patient_id"].nunique()),
       "tha": float((study["procedure"] == "THA").sum()), "revised": float(study["revised"].sum()),
       "mean_bmi": float(study["bmi"].mean())}
```

```{r}
#| include: false
check_agree(list(rows = nrow(study), patients = n_distinct(study$patient_id), tha = sum(study$procedure == "THA"),
                 revised = sum(study$revised), mean_bmi = mean(study$bmi, na.rm = TRUE)), reticulate::py$chk)
```

53 THAs and 67 TKAs. A left join keeps every workbook case even if the registry has no record of it; the next step checks that none is missing.

## 3 · Check the tidying {#integrity}

Before any analysis, check the tidy data against what you know must be true ([page 1](../foundations/01-tidy-data.qmd#integrity)). Write each check so that it **stops the script** when it fails: a check that only prints something is easy to scroll past.

::: {.panel-tabset group="language"}
## R

```{r}
# One row per case, and the 60 cases per site in the abstraction log
stopifnot(n_distinct(study$case_id) == nrow(study), all(count(study, site)$n == 60))

# Every value inside the range in the data dictionary (data/codebooks/abstraction_workbook_tidy.csv)
in_range <- function(x, low, high) all(is.na(x) | (x >= low & x <= high))
stopifnot(in_range(study$age, 40, 89), in_range(study$bmi, 18, 55), in_range(study$asa, 1, 4),
          in_range(study$los_days, 0, 30), in_range(study$prom_preop, 0, 100), in_range(study$prom_1yr, 0, 100))

# PROM dates inside the protocol's windows: pre-op within 30 days before surgery, 1 year at 10 to 14 months
stopifnot(in_range(as.numeric(study$surgery_date - study$prom_preop_date), 0, 30),
          in_range(as.numeric(study$prom_1yr_date - study$surgery_date), 300, 425))

# Every case found in the registry, and the workbook agrees with it on the surgery date,
# the procedure and the revisions (the red fills)
stopifnot(!anyNA(study$followup_years), all(study$surgery_date == study$surgery_date_registry),
          all(study$procedure == study$procedure_registry), all(study$revised == study$revised_registry))

# One row per patient: nobody in this sample had both hips or both knees replaced
stopifnot(n_distinct(study$patient_id) == nrow(study))

# What's missing: count it now, report it later
n_missing <- colSums(is.na(study))
n_missing[n_missing > 0]
```

## Python

```{python}
# One row per case, and the 60 cases per site in the abstraction log
assert study["case_id"].is_unique and (study["site"].value_counts() == 60).all()

# Every value inside the range in the data dictionary (data/codebooks/abstraction_workbook_tidy.csv)
def in_range(values, low, high):
    return bool(values.dropna().between(low, high).all())

assert in_range(study["age"], 40, 89) and in_range(study["bmi"], 18, 55) and in_range(study["asa"], 1, 4)
assert in_range(study["los_days"], 0, 30) and in_range(study["prom_preop"], 0, 100) and in_range(study["prom_1yr"], 0, 100)

# PROM dates inside the protocol's windows: pre-op within 30 days before surgery, 1 year at 10 to 14 months
assert in_range((study["surgery_date"] - study["prom_preop_date"]).dt.days, 0, 30)
assert in_range((study["prom_1yr_date"] - study["surgery_date"]).dt.days, 300, 425)

# Every case found in the registry, and the workbook agrees with it on the surgery date,
# the procedure and the revisions (the red fills)
assert study["followup_years"].notna().all() and (study["surgery_date"] == study["surgery_date_registry"]).all()
assert (study["procedure"] == study["procedure_registry"]).all() and (study["revised"] == study["revised_registry"]).all()

# One row per patient: nobody in this sample had both hips or both knees replaced
assert study["patient_id"].nunique() == len(study)

# What's missing: count it now, report it later
n_missing = study.isna().sum()
print(n_missing[n_missing > 0].to_string())
```
:::

```{python}
#| include: false
chk = {"bmi": float(n_missing["bmi"]), "asa": float(n_missing["asa"]), "preop": float(n_missing["prom_preop"]),
       "one_year": float(n_missing["prom_1yr"])}
```

```{r}
#| include: false
check_agree(list(bmi = n_missing[["bmi"]], asa = n_missing[["asa"]], preop = n_missing[["prom_preop"]],
                 one_year = n_missing[["prom_1yr"]]), reticulate::py$chk)
```

Every check passed, so the code ran on. The workbook agrees with the registry on the surgery date, the procedure and the revisions for all 120 cases. Checking one source against another catches what neither can alone: a red fill the abstractor forgot, or a study ID linked to the wrong registry record. The registry also shows that the 120 cases are 120 different patients, so each row is one patient. The missing values are what the abstractor couldn't find: 8 BMIs, 3 ASA classes, 3 lengths of stay, 5 pre-op scores and 22 one-year scores. The analysis has to say how it handles them.

## 4 · Table 1 {#table-1}

Table 1 describes the patients before surgery, by procedure ([page 2](../foundations/02-demographics.qmd#by-group)). In an observational study, show **standardized mean differences** (SMDs) instead of p-values ([page 2](../foundations/02-demographics.qmd#smd)): the question is how different the groups are, not whether the difference is "significant". Keep a Missing row for every variable that has missing values.

::: {.panel-tabset group="language"}
## R

```{r}
study <- study |> mutate(asa = factor(asa))   # ASA class is a category

table_one <- study |>
  select(procedure, age, sex, bmi, asa, diabetes, hypertension, sleep_apnea, site, prom_preop) |>
  tbl_summary(
    by           = procedure,
    statistic    = all_continuous() ~ "{mean} ({sd})",
    digits       = all_continuous() ~ 1,
    label        = list(age ~ "Age, years", sex ~ "Sex", bmi ~ "BMI, kg/m²", asa ~ "ASA class",
                        diabetes ~ "Diabetes", hypertension ~ "Hypertension", sleep_apnea ~ "Sleep apnea",
                        site ~ "Site", prom_preop ~ "Pre-op score (HOOS JR or KOOS JR), points"),
    missing_text = "Missing"
  ) |>                                                          # no Overall column: see below
  add_difference(test = everything() ~ "smd") |>
  modify_table_body(~ mutate(.x, estimate = abs(estimate))) |>  # absolute SMDs (page 2)
  modify_column_hide(columns = conf.low) |>                     # and not the SMDs' CIs
  remove_abbreviation("CI = Confidence Interval") |>
  modify_header(estimate = "**SMD**")
table_one
```

## Python

```{python}
study["asa"] = study["asa"].astype("Int64").astype("string")   # a category; blanks stay missing

labels = {"age": "Age, years", "sex": "Sex", "bmi": "BMI, kg/m²", "asa": "ASA class",
          "diabetes": "Diabetes", "hypertension": "Hypertension", "sleep_apnea": "Sleep apnea",
          "site": "Site", "prom_preop": "Pre-op score (HOOS JR or KOOS JR), points"}
table_one = TableOne(study, columns=list(labels), groupby="procedure", overall=False,   # no Overall column: see below
                     categorical=["sex", "asa", "diabetes", "hypertension", "sleep_apnea", "site"],
                     smd=True, missing=True, include_null=False, rename=labels)
print(table_one.tabulate(tablefmt="github"))
```
:::

```{python}
#| include: false
smd = table_one.smd_table.iloc[:, 0]
chk = {v: abs(float(smd[v])) for v in labels}
```

```{r}
#| include: false
smd_r <- table_one$table_body |> filter(row_type == "label")
# Continuous SMDs differ in the 3rd decimal: R's smd package divides the variances by n,
# tableone by n - 1 (page 2's R vs Python box).
check_agree(as.list(setNames(abs(smd_r$estimate), smd_r$variable)), reticulate::py$chk, tol = 0.005)
```

The groups differ most in sex: 72% of the THA patients were women against 49% of the TKA patients (SMD 0.47). Sleep apnea (0.31), ASA class (0.29) and the pre-op score (0.26) differ less; an SMD below 0.1 is usually read as a negligible difference. TKA patients started 2.9 points higher on average, which matters for an outcome measured as a change: a group that starts higher has less room to improve.

There's no Overall column. For most rows it would be harmless, but for the pre-op score it would average HOOS JR and KOOS JR scores together, which is the comparison Step 1 warned about.

::: {.callout-tip}
## 🔀 R vs Python: SMDs
The SMDs agree to two decimals. R's smd package divides the variances by *n*, tableone by *n* − 1, so continuous SMDs can differ in the third decimal ([page 2](../foundations/02-demographics.qmd#smd)). The hidden check allows for it. The two packages also subtract the groups in opposite orders, so their signs differ: the R table shows absolute values, as page 2 advises, and tableone's keep their signs.
:::

## 5 · Check the distributions {#distributions}

The primary outcome is the change from the pre-op score to the 1-year score, so it exists only for patients with both. Look at its shape within each group before choosing the test ([page 3](../foundations/03-distributions.qmd)), and count the patients at the top of the scale.

::: {.panel-tabset group="language"}
## R

```{r}
#| fig-height: 4.5
analysed <- study |>
  mutate(change = prom_1yr - prom_preop) |>   # the primary outcome
  filter(!is.na(change))                      # patients with both scores

ggplot(analysed, aes(change)) +
  geom_histogram(binwidth = 5, boundary = 0) +
  facet_wrap(~ procedure, ncol = 1) +
  labs(x = "Change in score, pre-op to 1 year (points)", y = "Patients")

change_summary <- analysed |>
  group_by(procedure) |>
  summarise(n = n(), mean = mean(change), sd = sd(change), at_ceiling = sum(prom_1yr == 100))
change_summary
```

## Python

```{python}
analysed = study.assign(change=study["prom_1yr"] - study["prom_preop"])   # the primary outcome
analysed = analysed[analysed["change"].notna()]                           # patients with both scores

fig, axes = plt.subplots(2, 1, figsize=(7, 4.5), sharex=True)
for ax, (procedure, rows) in zip(axes, analysed.groupby("procedure")):
    ax.hist(rows["change"], bins=range(0, 80, 5))
    ax.set_title(procedure)
    ax.set_ylabel("Patients")
axes[1].set_xlabel("Change in score, pre-op to 1 year (points)")
plt.tight_layout()
plt.show()

summary = analysed.groupby("procedure").agg(n=("change", "size"), mean=("change", "mean"),
                                            sd=("change", "std"), at_ceiling=("prom_1yr", lambda s: int((s == 100).sum())))
print(summary.round(1).to_string())
```
:::

```{python}
#| include: false
chk = {"n_tha": float(summary.loc["THA", "n"]), "n_tka": float(summary.loc["TKA", "n"]),
       "mean_tha": float(summary.loc["THA", "mean"]), "sd_tka": float(summary.loc["TKA", "sd"]),
       "ceiling_tha": float(summary.loc["THA", "at_ceiling"])}
```

```{r}
#| include: false
check_agree(list(n_tha = change_summary$n[1], n_tka = change_summary$n[2], mean_tha = change_summary$mean[1],
                 sd_tka = change_summary$sd[2], ceiling_tha = change_summary$at_ceiling[1]), reticulate::py$chk)
```

93 of the 120 patients have both scores: 46 THA and 47 TKA. Within each group the changes are roughly symmetric, with no long tail, and each group has more than 30 patients, so Welch's t test fits ([page 3](../foundations/03-distributions.qmd)).

One thing to note for the Discussion: 15 of the 46 THA patients (32.6%) scored the maximum, 100, at 1 year, against 7 of the 47 TKA patients (14.9%). A patient at the ceiling might have improved more if the scale allowed it, so the THA improvement, and the difference between the groups, may be underestimated. Exercise 1 checks the result with a rank-based test.

## 6 · The primary analysis {#primary-analysis}

The pre-specified analysis: Welch's t test on the change, with the mean difference, its 95% CI and Hedges' g ([page 6](../catalog/06-two-unpaired-groups.qmd#unpaired-t)).

::: {.panel-tabset group="language"}
## R

```{r}
primary <- t.test(change ~ procedure, data = analysed)   # Welch's t test is R's default
primary

primary_g <- effectsize::hedges_g(change ~ procedure, data = analysed)
primary_g
```

## Python

```{python}
tha = analysed.loc[analysed["procedure"] == "THA", "change"]
tka = analysed.loc[analysed["procedure"] == "TKA", "change"]

primary = stats.ttest_ind(tha, tka, equal_var=False)   # equal_var=False: Welch's t test
primary_ci = primary.confidence_interval()
print("t =", primary.statistic, " df =", primary.df, " p =", primary.pvalue)
print("difference =", tha.mean() - tka.mean(), " 95% CI", primary_ci.low, "to", primary_ci.high)

g = pg.compute_effsize(tha, tka, eftype="hedges")
g_ci = pg.compute_esci(stat=g, nx=len(tha), ny=len(tka), eftype="hedges")   # its 95% CI, to 2 decimals
print("Hedges' g =", g, " 95% CI", g_ci[0], "to", g_ci[1])
```
:::

```{python}
#| include: false
chk = {"t": float(primary.statistic), "df": float(primary.df), "p": float(primary.pvalue),
       "ci_low": float(primary_ci.low), "ci_high": float(primary_ci.high)}
chk_g = {"g": float(g)}
chk_g_ci = {"low": float(g_ci[0]), "high": float(g_ci[1])}
```

```{r}
#| include: false
check_agree(list(t = unname(primary$statistic), df = unname(primary$parameter), p = primary$p.value,
                 ci_low = primary$conf.int[1], ci_high = primary$conf.int[2]), reticulate::py$chk)
# effectsize corrects g for small samples exactly, pingouin with the usual approximation;
# with 93 patients they differ in the sixth decimal (see the R vs Python box)
check_agree(list(g = primary_g$Hedges_g), reticulate::py$chk_g, tol = 1e-5)
# pingouin gets the CI from an approximate formula and rounds it to 2 decimals; effectsize uses the
# noncentral t distribution. Here pingouin's CI is about 0.01 wider at each end, plus its rounding
# (see the R vs Python box), so this check only catches gross errors such as the wrong n.
check_agree(list(low = primary_g$CI_low, high = primary_g$CI_high), reticulate::py$chk_g_ci, tol = 0.02)
```

Reading it:
- **The difference:** THA patients improved by 42.4 points on average and TKA patients by 33.4, a difference of 9.0 points (95% CI 3.5 to 14.6). R puts THA first because it comes first alphabetically; the Python code subtracts in the same order.
- **p = 0.002:** if the procedures truly made no difference, a gap this large would be rare.
- **Hedges' g = 0.66** (95% CI 0.25 to 1.08): the difference in standard deviations. It's medium-to-large at the point estimate, but the CI runs from a small difference to a large one.

::: {.callout-tip}
## 🔀 R vs Python: Hedges' g
Hedges' g is Cohen's d shrunk slightly to correct for small samples. effectsize uses the exact correction, pingouin a close approximation, so with 93 patients they differ in the sixth decimal. For the CI, pingouin's `compute_esci()` uses an approximate formula and rounds to 2 decimals, so its CI, 0.24 to 1.09, is slightly wider than effectsize's 0.25 to 1.08, which comes from the noncentral t distribution. Say which you used; the manuscript quotes R's.
:::

## 7 · Revision: a survival analysis {#survival}

The secondary outcome is revision. First, how long were patients followed? Then, what proportion were revised over time? 18 patients died during follow-up without a revision, so death is a competing risk: the cumulative incidence gives the real proportion revised, where 1 − Kaplan-Meier would overstate it ([page 13](../survival/13-kaplan-meier.qmd#competing-risks)).

::: {.panel-tabset group="language"}
## R

```{r}
#| fig-height: 5
study <- study |>
  mutate(status = factor(event_status, levels = c(0, 1, 2),
                         labels = c("censored", "revision", "death")))   # censored must come first
table(study$status)

# median follow-up and IQR: the reverse Kaplan-Meier method
reverse_km <- survfit(Surv(followup_years, event_status == 0) ~ 1, data = study)
follow_up <- quantile(reverse_km, probs = c(0.25, 0.5, 0.75), conf.int = FALSE)
follow_up

# the cumulative incidence of revision, with death as a competing risk
cif <- survfit2(Surv(followup_years, status) ~ 1, data = study, conf.type = "log-log")
revision_rows <- tidy_survfit(cif, times = c(2, 5)) |>
  filter(outcome == "revision") |>
  select(time, estimate, conf.low, conf.high)
revision_rows

revision_plot <- cif |>
  ggcuminc(outcome = "revision") +
  add_confidence_interval() +
  add_risktable(risktable_stats = "n.risk") +                # the numbers at risk, under the curve
  scale_ggsurvfit(x_scales = list(breaks = seq(0, 10, by = 2))) +
  labs(x = "Years since surgery", y = "Cumulative incidence of revision")
revision_plot
```

## Python

```{python}
print(study["event_status"].value_counts().sort_index().to_string())   # 0 = censored, 1 = revision, 2 = death

# median follow-up and IQR: the reverse Kaplan-Meier method
reverse_km = KaplanMeierFitter().fit(study["followup_years"], event_observed=(study["event_status"] == 0))
follow_up = qth_survival_times([0.75, 0.5, 0.25], reverse_km.survival_function_).iloc[:, 0]   # 25th, 50th, 75th percentiles
print(follow_up.to_string())

# the cumulative incidence of revision, with death as a competing risk
cif = CumIncidenceRight(study["followup_years"], study["event_status"])
revision = pd.DataFrame({"cif": cif.cinc[0], "se": cif.cinc_se[0]}, index=cif.times)   # event type 1

z = norm.ppf(0.975)
revision_rows = revision.loc[[revision.index[revision.index <= years][-1] for years in [2, 5]]]   # last step before each time
spread = np.exp(z * revision_rows["se"] / (revision_rows["cif"] * np.log(revision_rows["cif"])))  # the log-log CI, as in R
revision_rows = revision_rows.assign(time=[2, 5], low=revision_rows["cif"] ** (1 / spread),
                                     high=revision_rows["cif"] ** spread)
print(revision_rows[["time", "cif", "low", "high"]].round(3).to_string(index=False))

fig, ax = plt.subplots(figsize=(7, 4))
ax.step(revision.index, revision["cif"], where="post")
ax.set_xlabel("Years since surgery")
ax.set_ylabel("Cumulative incidence of revision")
plt.tight_layout()
plt.show()
```
:::

```{python}
#| include: false
chk = {"q1": float(follow_up.iloc[0]), "median": float(follow_up.iloc[1]), "q3": float(follow_up.iloc[2]),
       "cif_2": float(revision_rows["cif"].iloc[0]), "cif_5": float(revision_rows["cif"].iloc[1]),
       "low_5": float(revision_rows["low"].iloc[1]), "high_5": float(revision_rows["high"].iloc[1])}
```

```{r}
#| include: false
check_agree(list(q1 = follow_up[[1]], median = follow_up[[2]], q3 = follow_up[[3]],
                 cif_2 = revision_rows$estimate[1], cif_5 = revision_rows$estimate[2],
                 low_5 = revision_rows$conf.low[2], high_5 = revision_rows$conf.high[2]),
            reticulate::py$chk)
```

The median follow-up was 6.05 years (IQR 4.16 to 8.27). 18 patients were revised and 18 died without a revision. The cumulative incidence of revision was 3.4% (95% CI 1.1% to 7.8%) at 2 years and 12.6% (95% CI 7.0% to 19.8%) at 5 years. With 18 revisions, the CIs are wide: this cohort can describe revision, not compare it between procedures. The numbers at risk under the R plot show how quickly the data thin out: 17 patients were still followed at 8 years and 5 at 10, so the end of the curve rests on very few patients. Publish them with the curve ([page 13](../survival/13-kaplan-meier.qmd#kaplan-meier)).

::: {.callout-tip}
## 🔀 R vs Python: the cumulative incidence
R's `survfit()` estimates it whenever the status has three levels. statsmodels' `CumIncidenceRight()` gives the same estimates and their standard errors, but no CI, so the Python code builds R's log-log CI by hand ([page 13](../survival/13-kaplan-meier.qmd#competing-risks)).
:::

## 8 · The manuscript {#manuscript}

Everything from here to the exercises is the manuscript, and it's what the Word version contains. Every number in it comes from the code above, and a hidden check stops the render if the text and the code ever disagree.

::: {.callout-note}
## 💡 What "done" looks like
- The question, outcome and analysis were written down before the analysis ran.
- The tidying is checked, and a failed check stops the script.
- Missing data are counted and reported, not dropped silently.
- Every estimate has a 95% CI, and the Methods name every test.
- The manuscript is rendered from the same file as the analysis. If the data change, Table 1 and Figure 1 update themselves, and the hidden check stops the render until every number in the text is updated by hand.
:::

The manuscript stops at the Results. The Discussion is the author's to write; for this study it would need to cover the ceiling effect at 1 year, the patients missing from the analysis, the two questionnaires, and whether a 9-point difference matters to patients.
:::

## Aim {#aim}

To compare the improvement in joint-specific patient-reported scores from before surgery to 1 year after primary total hip arthroplasty (THA) and total knee arthroplasty (TKA), and to describe the cumulative incidence of revision.

## Methods {#methods}

**Patients and data.** We studied 120 primary THAs and TKAs performed at two sites (60 at each) between January 2015 and November 2024. Patients' characteristics, the procedure and patient-reported scores were abstracted from the medical record into a standard sheet. Follow-up for revision and death came from the joint registry, linked by study ID.

**Outcomes.** The primary outcome was the change in the joint-specific score from before surgery to 1 year: the HOOS JR after THA and the KOOS JR after TKA, each scored from 0 to 100 (higher is better). Because the two are different instruments, the comparison is of points on each joint's own scale. The secondary outcome was revision of any component.

**Statistical analysis.** Continuous variables are reported as mean (SD) and categorical variables as n (%); differences between procedures at baseline are described with standardized mean differences. The change in score was compared between procedures with Welch's t test in patients with both scores, and summarized as the mean difference with its 95% CI and Hedges' g. For revision, time was measured from the date of surgery to revision, death without revision (a competing event) or the end of registry follow-up (censored). Median follow-up was estimated with the reverse Kaplan-Meier method. The cumulative incidence of revision was estimated with the Aalen-Johansen method, treating death as a competing risk. Tests were two-sided. Analyses used R 4.6 and Python 3.13.

## Results {#results}

Of the 120 patients, 53 (44.2%) had a THA and 67 (55.8%) a TKA (@tbl-baseline). Women made up 38 of 53 THA patients (71.7%) and 33 of 67 TKA patients (49.3%; SMD 0.47). The mean pre-op score was 48.1 points (SD 11.5) in THA patients and 51.0 points (SD 11.8) in TKA patients (SMD 0.26).

Both scores were available for 93 patients (77.5%): 46 of 53 after THA (86.8%) and 47 of 67 after TKA (70.1%). The mean change in score was 42.4 points (SD 13.0) after THA and 33.4 points (SD 13.9) after TKA (difference 9.0 points, 95% CI 3.5 to 14.6; Hedges' g 0.66, 95% CI 0.25 to 1.08; p = 0.002). At 1 year, 15 of 46 THA patients (32.6%) and 7 of 47 TKA patients (14.9%) had the maximum score.

Over a median follow-up of 6.05 years (IQR 4.16 to 8.27), 18 patients had a revision and 18 died without one. The cumulative incidence of revision was 3.4% (95% CI 1.1% to 7.8%) at 2 years and 12.6% (95% CI 7.0% to 19.8%) at 5 years (@fig-revision).

```{r}
#| echo: false
#| label: tbl-baseline
#| tbl-cap: "Patients' characteristics before surgery, by procedure. Mean (SD) or n (%); SMD, absolute standardized mean difference."
as_flex_table(table_one)
```

```{r}
#| echo: false
#| label: fig-revision
#| fig-cap: "Cumulative incidence of revision, with death as a competing risk. The shaded band is the 95% CI."
#| fig-height: 5
revision_plot
```

```{r}
#| include: false
# Prose guard: numbers quoted in the text above. If the data change, update the text.
smd_of <- function(v) round(abs(smd_r$estimate[smd_r$variable == v]), 2)
sex_counts <- table(study$procedure, study$sex)
ceiling_counts <- analysed |> group_by(procedure) |> summarise(n = sum(prom_1yr == 100), pct = 100 * mean(prom_1yr == 100))
both_counts <- study |> group_by(procedure) |> summarise(n = n(), both = sum(!is.na(prom_1yr - prom_preop)))
pre_means <- study |> group_by(procedure) |> summarise(mean = mean(prom_preop, na.rm = TRUE), sd = sd(prom_preop, na.rm = TRUE))
ancova <- lm(prom_1yr ~ prom_preop + procedure, data = analysed)
ancova_ci <- confint(ancova)["procedureTKA", ]
rank_check <- wilcox.test(change ~ procedure, data = analysed, exact = FALSE, conf.int = TRUE)
missing_test <- fisher.test(table(study$procedure, is.na(study$prom_1yr - study$prom_preop)))
stopifnot(
  nrow(study) == 120, all(table(study$site) == 60), ncol(tidy) == 18,
  format(min(study$surgery_date), "%B %Y") == "January 2015", format(max(study$surgery_date), "%B %Y") == "November 2024",
  as.vector(table(study$procedure)) == c(53, 67), round(100 * 53 / 120, 1) == 44.2, round(100 * 67 / 120, 1) == 55.8,
  n_missing[c("bmi", "asa", "los_days", "prom_preop", "prom_1yr")] == c(8, 3, 3, 5, 22),
  sex_counts["THA", "Female"] == 38, sex_counts["TKA", "Female"] == 33,
  round(100 * 38 / 53, 1) == 71.7, round(100 * 33 / 67, 1) == 49.3, round(100 * 38 / 53) == 72, round(100 * 33 / 67) == 49,
  smd_of("sex") == 0.47, smd_of("sleep_apnea") == 0.31, smd_of("asa") == 0.29, smd_of("prom_preop") == 0.26,
  round(pre_means$mean, 1) == c(48.1, 51.0), round(pre_means$sd, 1) == c(11.5, 11.8), round(round(pre_means$mean[2], 1) - round(pre_means$mean[1], 1), 1) == 2.9,
  nrow(analysed) == 93, round(100 * 93 / 120, 1) == 77.5,
  both_counts$both == c(46, 47), round(100 * both_counts$both / both_counts$n, 1) == c(86.8, 70.1),
  round(change_summary$mean, 1) == c(42.4, 33.4), round(change_summary$sd, 1) == c(13.0, 13.9),
  round(diff(rev(primary$estimate)), 1) == 9.0, round(primary$conf.int, 1) == c(3.5, 14.6),
  round(primary$p.value, 3) == 0.002,
  round(primary_g$Hedges_g, 2) == 0.66, round(c(primary_g$CI_low, primary_g$CI_high), 2) == c(0.25, 1.08),
  round(reticulate::py$g_ci, 2) == c(0.24, 1.09),
  ceiling_counts$n == c(15, 7), round(ceiling_counts$pct, 1) == c(32.6, 14.9),
  table(study$event_status) == c(84, 18, 18),
  sum(study$followup_years >= 8) == 17, sum(study$followup_years >= 10) == 5,
  round(unname(follow_up), 2) == c(4.16, 6.05, 8.27),
  round(100 * revision_rows$estimate, 1) == c(3.4, 12.6),
  round(100 * revision_rows$conf.low, 1) == c(1.1, 7.0), round(100 * revision_rows$conf.high, 1) == c(7.8, 19.8),
  R.version$major == "4", startsWith(R.version$minor, "6"),
  reticulate::py_config()$version == "3.13",
  # exercise solutions (recomputed here: the guard runs before the solution chunks)
  round(unname(rank_check$estimate), 1) == 9.5, round(rank_check$conf.int, 1) == c(3.6, 15.2),
  round(rank_check$p.value, 3) == 0.003,
  round(coef(ancova)[["procedureTKA"]], 1) == -7.4, round(ancova_ci, 1) == c(-11.9, -2.9),
  round(summary(ancova)$coefficients["procedureTKA", "Pr(>|t|)"], 3) == 0.002,
  table(study$procedure, is.na(study$prom_1yr - study$prom_preop))[, "TRUE"] == c(7, 20),
  round(100 * c(7 / 53, 20 / 67), 1) == c(13.2, 29.9),
  tapply(is.na(study$prom_preop), study$procedure, sum) == c(0, 5),
  tapply(is.na(study$prom_1yr), study$procedure, sum) == c(7, 15), round(100 * 15 / 67, 1) == 22.4,
  round(fisher.test(table(study$procedure, is.na(study$prom_1yr)))$p.value, 3) == 0.239,
  round(missing_test$p.value, 3) == 0.046
)
```

::: {.content-visible when-format="html"}
## Exercises {#exercises}

The solutions use the packages, data and objects made in the steps above, so run the page from the top first.

**1.** Fifteen THA patients scored the maximum at 1 year. Check the primary result with a test that uses ranks instead of means: the Mann-Whitney test and the Hodges-Lehmann difference ([page 6](../catalog/06-two-unpaired-groups.qmd#mann-whitney)). Does the conclusion change?

::: {.callout-tip collapse="true"}
## Solution

::: {.panel-tabset group="language"}
## R

```{r}
wilcox.test(change ~ procedure, data = analysed, exact = FALSE, conf.int = TRUE)
```

## Python

```{python}
rank_check = stats.mannwhitneyu(tha, tka, alternative="two-sided", method="asymptotic", use_continuity=True)
print(rank_check.statistic, rank_check.pvalue)
print(np.median(np.subtract.outer(tha.to_numpy(), tka.to_numpy())))   # the Hodges-Lehmann difference
```
:::

> **Results:** The Hodges-Lehmann difference in the change was 9.5 points (95% CI 3.6 to 15.2; p = 0.003).

The conclusion doesn't change: 9.5 points is close to the 9.0 from the t test. Python gives the same estimate and p-value; its CI needs a bootstrap ([page 6](../catalog/06-two-unpaired-groups.qmd#mann-whitney)).
:::

**2.** A reviewer writes: "TKA patients started with higher scores, so they had less room to improve. Adjust for the pre-op score." Which analysis answers this, and what does it show?

::: {.callout-tip collapse="true"}
## Solution

Regress the 1-year score on the procedure and the pre-op score ([page 12](../catalog/12-predict-from-several.qmd#multiple-linear-regression)). The procedure's coefficient is the difference at 1 year between patients who started with the same score.

::: {.panel-tabset group="language"}
## R

```{r}
adjusted <- lm(prom_1yr ~ prom_preop + procedure, data = analysed)
summary(adjusted)$coefficients
confint(adjusted)
```

## Python

```{python}
adjusted = smf.ols("prom_1yr ~ prom_preop + procedure", data=analysed).fit()
print(adjusted.params.round(3).to_string())
print(adjusted.conf_int().round(3).to_string())
print(adjusted.pvalues.round(3).to_string())
```
:::

> **Results:** Adjusted for the pre-op score, the difference in the 1-year score between TKA and THA was −7.4 points (95% CI −11.9 to −2.9; p = 0.002).

The difference shrinks a little from the unadjusted 9.0 points but doesn't go away, so the higher starting point explains only part of it.
:::

**3.** 27 patients have no change score, so the primary analysis leaves them out. Were they left out more often after one procedure, and which score was missing? Test it and write the Results sentence.

::: {.callout-tip collapse="true"}
## Solution

::: {.panel-tabset group="language"}
## R

```{r}
no_change_score <- is.na(study$prom_1yr - study$prom_preop)
table(study$procedure, no_change_score)
fisher.test(table(study$procedure, no_change_score))$p.value

# which score was missing?
study |>
  group_by(procedure) |>
  summarise(no_preop = sum(is.na(prom_preop)), no_1yr = sum(is.na(prom_1yr)))
fisher.test(table(study$procedure, is.na(study$prom_1yr)))$p.value   # the 1-year score alone
```

## Python

```{python}
no_change_score = (study["prom_1yr"] - study["prom_preop"]).isna().rename("no_change_score")
missing_table = pd.crosstab(study["procedure"], no_change_score)
print(missing_table.to_string())
print(stats.fisher_exact(missing_table).pvalue)

# which score was missing?
print(study[["prom_preop", "prom_1yr"]].isna().groupby(study["procedure"]).sum().to_string())
print(stats.fisher_exact(pd.crosstab(study["procedure"], study["prom_1yr"].isna())).pvalue)   # the 1-year score alone
```
:::

> **Results:** The change score was missing for 7 of 53 THA patients (13.2%) and 20 of 67 TKA patients (29.9%; Fisher's exact test p = 0.046): all 7 THA patients and 15 TKA patients had no 1-year score, and the other 5 TKA patients had no pre-op score. The 1-year score alone was missing for 7 THA patients (13.2%) and 15 TKA patients (22.4%; p = 0.239).

If the patients left out fared differently from those analysed, the comparison is biased; the Discussion should say so.
:::
:::
