---
title: "2 · Demographics (Table 1)"
description: "Build a publication-ready Table 1 for one group or several: choosing summaries, comparing groups, standardized mean differences, missing data, and exporting to Word."
engine: knitr
---

```{r}
#| include: false
source("R/check_agree.R")
```

```{r}
#| echo: false
#| output: asis
# Stamp the data version into the page; tests/site/test_freshness.py checks it.
cat(sprintf("<!-- data-checksum: %s -->", unname(tools::md5sum("data/CHECKSUMS.md5"))))
```

Almost every clinical paper starts with **Table 1**: who was studied. Reviewers read it first, to decide whether your patients look like theirs and whether your groups were comparable to begin with. This page builds one from scratch.

::: {.callout-note}
## 💡 In plain language
Table 1 *describes* your patients. It doesn't answer your research question. Its job is to let a reader judge who your results apply to, and whether your groups started out alike.
:::

## Load the data and pick the variables {#setup}

We'll describe the practice cohort: 600 primary hip and knee replacements (THA and TKA). Table 1 usually holds the characteristics measured *before* surgery.

::: {.panel-tabset group="language"}
## R

```{r}
#| warning: false
library(tidyverse)
library(gtsummary)

cohort <- read_csv("data/cohort.csv", show_col_types = FALSE) |>
  mutate(asa = factor(asa))   # ASA class is a category, not a number

table1_vars <- c("age", "sex", "bmi", "asa", "diabetes", "smoker", "los_days")

labels <- list(age ~ "Age, years", sex ~ "Sex", bmi ~ "BMI, kg/m²", asa ~ "ASA class",
               diabetes ~ "Diabetes", smoker ~ "Smoking status",
               los_days ~ "Length of stay, days")
```

## Python

```{python}
import pandas as pd
from tableone import TableOne

cohort = pd.read_csv("data/cohort.csv")
cohort["asa"] = cohort["asa"].astype(str)   # ASA class is a category, not a number

table1_vars = ["age", "sex", "bmi", "asa", "diabetes", "smoker", "los_days"]
categorical = ["sex", "asa", "diabetes", "smoker"]

labels = {"age": "Age, years", "sex": "Sex", "bmi": "BMI, kg/m²", "asa": "ASA class",
          "diabetes": "Diabetes", "smoker": "Smoking status",
          "los_days": "Length of stay, days"}
```
:::

::: {.callout-warning}
## ⚠️ Watch out: one row per procedure, not per patient
The cohort has one row per *procedure*. 80 patients had both sides done, so 600 procedures come from 520 patients. Say which one your table counts, for example in a footnote: "600 procedures in 520 patients". The two procedures of a bilateral patient are not independent (same person, same age, sex and health), so tests that treat every row as a separate patient are slightly overconfident. [Distributions & choosing a test](03-distributions.qmd#paired) explains the options.
:::

## One group {#one-group}

::: {.panel-tabset group="language"}
## R

```{r}
table_one <- cohort |>
  select(all_of(table1_vars)) |>
  tbl_summary(
    type      = list(los_days ~ "continuous"),          # few distinct values, but still a number
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     los_days ~ "{median} ({p25}–{p75})"),
    digits    = list(all_continuous() ~ 1, los_days ~ 0),   # 65.5 (9.4); whole days
    label     = labels
  )
table_one
```

## Python

```{python}
table_one = TableOne(
    cohort,
    columns=table1_vars,
    categorical=categorical,
    nonnormal=["los_days"],      # report median [Q1, Q3] instead of mean (SD)
    rename=labels,
)
print(table_one.tabulate(tablefmt="github"))
```
:::

```{python}
#| include: false
chk = {"n": float(len(cohort)), "mean_age": float(cohort["age"].mean()),
       "sd_age": float(cohort["age"].std()), "female": float((cohort["sex"] == "Female").sum()),
       "median_los": float(cohort["los_days"].median())}
```

```{r}
#| include: false
check_agree(list(n = nrow(cohort), mean_age = mean(cohort$age), sd_age = sd(cohort$age),
                 female = sum(cohort$sex == "Female"), median_los = median(cohort$los_days)),
            reticulate::py$chk)
```

::: {.callout-warning}
## ⚠️ Watch out: tell the software what each variable is
Software guesses variable types, and the guesses are often wrong:

- **ASA class** looks like a number, so it gets averaged ("mean ASA 2.4"). It's a category.
- **Length of stay** has only a few distinct values, so gtsummary treats it as a category and lists "0 days, 1 day, 2 days…". It's a count.
- **0/1 columns** such as `diabetes`: gtsummary shows a single row (how many are 1, i.e. yes), while tableone shows both the 0 and the 1 row. In a paper, show just the "yes" row.

Set the type of every variable yourself (`type =` in R, `categorical =` in Python).
:::

## Mean (SD) or median (IQR)? {#which-summary}

- **Roughly symmetric** continuous variables (age, BMI): mean (SD).
- **Skewed or bounded** variables (length of stay, most PROMs): median (IQR). One patient with a 30-day stay drags the mean far more than the median.
- **Categories**: n (%).

How do you know whether a variable is skewed? Look at it: [Distributions & choosing a test](03-distributions.qmd) shows how. Length of stay is strongly right-skewed, which is why it gets a median.

::: {.callout-tip}
## 🔀 R vs Python: quartiles
gtsummary and tableone use different formulas for quartiles (in R's terms, `quantile()` types 2 and 7). With whole-number data the results can differ. With whole-number data like length of stay you can see the difference: the THA upper quartile is 2 days in R and 1.8 in Python. Neither is wrong. Say in your Methods which software you used.
:::

## Comparing groups {#by-group}

Most studies compare groups. Here: THA vs TKA.

::: {.panel-tabset group="language"}
## R

```{r}
by_procedure <- cohort |>
  select(procedure, all_of(table1_vars)) |>
  tbl_summary(
    by        = procedure,
    type      = list(los_days ~ "continuous"),
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     los_days ~ "{median} ({p25}–{p75})"),
    digits    = list(all_continuous() ~ 1, los_days ~ 0),
    label     = labels
  ) |>
  add_overall() |>
  add_p(test = list(all_continuous() ~ "t.test",        # Welch's t test
                    los_days ~ "kruskal.test",           # rank-based, for the skewed variable
                    all_categorical() ~ "chisq.test"),   # chi-square
        pvalue_fun = label_style_pvalue(digits = 3))     # p to three decimals
by_procedure
```

## Python

```{python}
by_procedure = TableOne(
    cohort,
    columns=table1_vars,
    categorical=categorical,
    nonnormal=["los_days"],
    groupby="procedure",
    pval=True,               # Welch's t test, Kruskal-Wallis, chi-square
    htest_name=True,         # show which test produced each p-value
    rename=labels,
)
print(by_procedure.tabulate(tablefmt="github"))
```
:::

```{python}
#| include: false
p_values = by_procedure.htest_table["P-Value"]
chk = {v: float(p_values[v]) for v in table1_vars}
```

```{r}
#| include: false
p_r <- by_procedure$table_body |> filter(row_type == "label")
check_agree(as.list(setNames(p_r$p.value, p_r$variable)), reticulate::py$chk)
```

Each p-value comes from a test you'll meet in the catalog: [Welch's t test](../catalog/06-two-unpaired-groups.qmd#unpaired-t) for the symmetric continuous variables, a [rank-based test](../catalog/06-two-unpaired-groups.qmd#mann-whitney) for length of stay (Kruskal-Wallis with two groups is the Mann-Whitney test), and the [chi-square test](../catalog/06-two-unpaired-groups.qmd#fisher-chi-square) for the categories.

## p-values or standardized mean differences? {#smd}

Look at sex in the table above: 66% of THA patients are women against 50% of TKA patients, p < 0.001. Is that a real difference between hip and knee patients?

In this practice dataset we know the answer, because we wrote the program that made it: sex was drawn **completely independently** of procedure. The difference is pure chance. (In real arthroplasty cohorts, TKA patients are usually *more* often female.) That's the first problem with p-values in Table 1: across a long table, chance differences turn up.

The second problem is that a p-value mixes up *how big* a difference is with *how many patients* you have. With 6,000 patients a trivial difference gets p < 0.001. With 30, a large one doesn't.

The **standardized mean difference (SMD)** measures only the size of the imbalance, in standard-deviation units, whatever the sample size. A common rule of thumb: |SMD| below 0.1 is negligible.

::: {.panel-tabset group="language"}
## R

```{r}
balance <- cohort |>
  select(procedure, all_of(table1_vars)) |>
  tbl_summary(
    by        = procedure,
    type      = list(los_days ~ "continuous"),
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     los_days ~ "{median} ({p25}–{p75})"),
    digits    = list(all_continuous() ~ 1, los_days ~ 0),
    label     = labels
  ) |>
  add_difference(test = everything() ~ "smd")
balance
```

## Python

```{python}
balance = TableOne(
    cohort,
    columns=table1_vars,
    categorical=categorical,
    nonnormal=["los_days"],
    groupby="procedure",
    smd=True,
    rename=labels,
)
print(balance.smd_table)
```
:::

```{python}
#| include: false
smd = balance.smd_table.iloc[:, 0]
chk = {v: abs(float(smd[v])) for v in table1_vars}
```

```{r}
#| include: false
smd_r <- balance$table_body |> filter(row_type == "label")
# Continuous SMDs differ in the 3rd decimal (variance divided by n in R's smd
# package, n - 1 in tableone), documented in the R vs Python box below.
check_agree(as.list(setNames(abs(smd_r$estimate), smd_r$variable)), reticulate::py$chk, tol = 0.005)
```

Sex (SMD 0.31) and BMI (0.33) are the clear imbalances. Age (0.10), ASA class (0.12) and smoking (0.15) are just over 0.1: worth a sentence, not alarm. Diabetes and length of stay are negligible.

::: {.callout-tip}
## 🔀 R vs Python: SMDs differ slightly
- **Sign.** gtsummary subtracts TKA from THA; tableone subtracts THA from TKA. Report the absolute value, or say which group was subtracted from which.
- **Fourth decimal.** For continuous variables, R's `smd` package divides the variances by *n*, while tableone divides by *n − 1* (the usual textbook formula). Age comes out as 0.1048 in R and 0.1046 in Python. Categorical SMDs match exactly.
:::

**Which should you report?**

- **Matched or weighted cohorts** (propensity-score studies, such as TJS's age-80 matched study): SMDs only. The matching made the groups alike on purpose, so a test of "did they differ?" makes no sense.
- **Other observational comparisons**: many journals still expect p-values. Give SMDs alongside, and discuss differences by their size, not their p-value.

## Missing data {#missing}

The tidy version of the messy workbook from [page 1](01-tidy-data.qmd) has a few BMIs, ASA classes and lengths of stay that the abstractor couldn't find. Missing values must show up in Table 1. Never drop them silently.

::: {.panel-tabset group="language"}
## R

```{r}
abstraction <- read_csv("data/answer-keys/abstraction_workbook_tidy.csv", show_col_types = FALSE) |>
  mutate(asa = factor(asa))

abstraction |>
  select(site, age, bmi, asa, los_days) |>
  tbl_summary(by = site, type = list(los_days ~ "continuous"),
              statistic = list(all_continuous() ~ "{mean} ({sd})",
                               los_days ~ "{median} ({p25}–{p75})"),
              digits = list(all_continuous() ~ 1, los_days ~ 0),
              missing_text = "Missing")
```

## Python

```{python}
abstraction = pd.read_csv("data/answer-keys/abstraction_workbook_tidy.csv")
abstraction["asa"] = abstraction["asa"].astype("Int64").astype("string")   # a category; blanks stay missing

missing_table = TableOne(abstraction, columns=["age", "bmi", "asa", "los_days"],
                         categorical=["asa"], nonnormal=["los_days"],
                         groupby="site", missing=True,
                         include_null=False)   # count blanks as missing, not as a "None" category
print(missing_table.tabulate(tablefmt="github"))
```
:::

```{python}
#| include: false
chk = {"missing_bmi": float(abstraction["bmi"].isna().sum()),
       "missing_asa": float(abstraction["asa"].isna().sum())}
```

```{r}
#| include: false
check_agree(list(missing_bmi = sum(is.na(abstraction$bmi)),
                 missing_asa = sum(is.na(abstraction$asa))), reticulate::py$chk)
```

R adds a "Missing" row under each variable that has any. Python adds a "Missing" column. Either way, the percentages are of the patients with a value, and the reader can see how many were left out.

## Exporting to Word {#export}

Journals want Table 1 in Word. Save it into your `scratch/` folder (git ignores it), open it, and paste it into the manuscript.

::: {.panel-tabset group="language"}
## R

```{r}
by_procedure |>
  as_flex_table() |>
  flextable::save_as_docx(path = "scratch/table1.docx")
```

## Python

```{python}
from docx import Document

table = by_procedure.tableone.reset_index()
headers = [" ".join(str(part) for part in col if part).strip() for col in table.columns]

doc = Document()
grid = doc.add_table(rows=1, cols=len(headers))
for cell, header in zip(grid.rows[0].cells, headers):
    cell.text = header
for _, row in table.iterrows():
    for cell, value in zip(grid.add_row().cells, row):
        cell.text = "" if pd.isna(value) else str(value)
doc.save("scratch/table1-python.docx")
```
:::

In Word, tidy the formatting to the journal's style: usually no vertical lines, and a footnote that says which statistics are shown ("mean (SD); median (IQR); n (%)") and which tests produced the p-values.

```{r}
#| include: false
# Prose guard: numbers quoted in the text above. If the data change, update the text.
pct <- function(x) round(100 * mean(x))
site_table <- cohort |>
  select(site, all_of(table1_vars)) |>
  tbl_summary(by = site, type = list(los_days ~ "continuous")) |>
  add_difference(test = everything() ~ "smd")
site_rows <- site_table$table_body |> filter(row_type == "label")
site_smd <- setNames(abs(site_rows$estimate), site_rows$variable)
stopifnot(
  nrow(cohort) == 600, n_distinct(cohort$patient_id) == 520,
  sum(table(cohort$patient_id) == 2) == 80,
  sum(cohort$procedure == "THA") == 270, sum(cohort$procedure == "TKA") == 330,
  pct(cohort$sex[cohort$procedure == "THA"] == "Female") == 66,
  pct(cohort$sex[cohort$procedure == "TKA"] == "Female") == 50,
  round(mean(cohort$age), 1) == 65.5, round(sd(cohort$age), 1) == 9.4,
  sum(cohort$sex == "Female") == 343, pct(cohort$sex == "Female") == 57,
  median(cohort$los_days) == 1,
  # what the libraries actually display (Python side checked in the next chunk)
  by_procedure$table_body$stat_1[by_procedure$table_body$variable == "los_days" &
                                   by_procedure$table_body$row_type == "label"] == "1 (0–2)",
  by_procedure$table_body$stat_0[by_procedure$table_body$variable == "los_days" &
                                   by_procedure$table_body$row_type == "label"] == "1 (0–1)",
  by_procedure$table_body$stat_0[by_procedure$table_body$variable == "age" &
                                   by_procedure$table_body$row_type == "label"] == "65.5 (9.4)",
  round(abs(smd_r$estimate[smd_r$variable == "age"]), 2) == 0.10,
  round(abs(smd_r$estimate[smd_r$variable == "asa"]), 2) == 0.12,
  round(abs(smd_r$estimate[smd_r$variable == "smoker"]), 2) == 0.15,
  abs(smd_r$estimate[smd_r$variable == "diabetes"]) < 0.1,
  abs(smd_r$estimate[smd_r$variable == "los_days"]) < 0.1,
  abs(site_smd[["smoker"]] - 0.23) < 0.005, abs(site_smd[["los_days"]] - 0.225) < 0.005,
  all(site_smd[setdiff(names(site_smd), c("smoker", "los_days"))] < 0.11),
  round(abs(smd_r$estimate[smd_r$variable == "sex"]), 2) == 0.31,
  round(abs(smd_r$estimate[smd_r$variable == "bmi"]), 2) == 0.33
)
```

```{python}
#| include: false
# Prose guard (Python side): the quartile tableone displays.
assert by_procedure.tableone.loc[("Length of stay, days, median [Q1,Q3]", ""),
                                 ("Grouped by procedure", "THA")] == "1.0 [0.0,1.8]"
```

## Exercises {#exercises}

**1.** Make a Table 1 comparing **Site A with Site B** instead of THA with TKA. Which variable differs most between sites, by SMD?

::: {.callout-tip collapse="true"}
## Solution

::: {.panel-tabset group="language"}
## R

```{r}
cohort |>
  select(site, all_of(table1_vars)) |>
  tbl_summary(by = site, type = list(los_days ~ "continuous"),
              statistic = list(all_continuous() ~ "{mean} ({sd})",
                               los_days ~ "{median} ({p25}–{p75})"),
              digits = list(all_continuous() ~ 1, los_days ~ 0),
              label = labels) |>
  add_difference(test = everything() ~ "smd")
```

## Python

```{python}
by_site = TableOne(cohort, columns=table1_vars, categorical=categorical,
                   nonnormal=["los_days"], groupby="site", smd=True, rename=labels)
print(by_site.smd_table)
```
:::

Smoking status (0.23) and length of stay (0.22) differ most between sites, and their 95% CIs exclude zero. Yet site was assigned completely at random in this dataset, independent of everything else. With seven variables, a couple of chance imbalances above 0.1 are to be expected: judge each SMD by its size and by whether it could plausibly matter, not by whether it is "significant".
:::

**2.** A fellow is comparing patients aged 80 or older with younger patients, matched on sex, BMI and ASA class. Should the baseline table show p-values or SMDs? Why?

::: {.callout-tip collapse="true"}
## Solution
SMDs. The matching was designed to make the groups alike, so testing whether they differ answers a question nobody asked, and with matched data the p-value mostly reflects sample size. SMDs show how well the matching worked: aim for |SMD| below 0.1 on every matching variable.
:::

**3.** Write the first sentence of the Results describing this cohort, using the reporting conventions from [How to use this site](../getting-started/using-this-site.qmd).

::: {.callout-tip collapse="true"}
## Solution
"We included 600 primary arthroplasties (270 THA and 330 TKA) in 520 patients. Mean age was 65.5 (SD 9.4) years and 343 procedures (57%) were in women. Median length of stay was 1 (IQR 0–1) day."

Check every number against your own table: mean and SD for age, n (%) for sex, median (IQR) for length of stay.
:::
